\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-23 k-planar.tex
Pfffft

\citet{DJMMUW20} introduced the study of partitions with bounded layered width such that the quotient has some additional desirable property, like small treewidth. Dujmovi\'c~et~al.\ define a class $\mathcal{G}$ of graphs to \defin{admit bounded layered partitions} if there exist $t,\ell\in\mathbb{N}$ such that every graph $G\in \mathcal{G}$ has an $H$-partition of layered width at most $\ell$ for some graph $H=H(G)$ of treewidth at most $t$.

@@

The following result is the main technical contribution of the paper. Loosely speaking, it shows that if a graph $G$ admits bounded layered partitions, then so too does $G^\SS$ for every shortcut system $\SS$ of $G$.

@@

Recall that the \defin{$k$-th power} of a graph $G$ is the graph $G^k$ with vertex set $V(G^k):=V(G)$, where $vw\in E(G^k)$ if and only if $\dist_G(v,w)\leq k$. If $G$ is planar with maximum degree $\Delta$, then $G^k$ is $2k(k+1)\Delta^{k}$-planar by \cref{DrawG}. Thus we can immediately conclude that bounded powers of planar graphs of bounded degree admit bounded layered partitions. However, the bounds we obtain are improved by the following lemma that constructs a shortcut system directly.

@@

@@

\section{Framed Graphs, $1$-Planar Graphs, and Map Graphs}
\label{FramedSection}

@@

\note{PM}{Check David's commit: a0e20aff83d12c74127331854ce9fb75e002efe3}

\note{DW}{This subsection needs a clean-up; define $(g,d)$-framed graph.}

\begin{thm}
\label{NewMapGraph}
Every $d$-map graph is contained in $H \boxtimes P \boxtimes K_{ d + 3\floor{d/2} -3 }$ for some planar graph $H$ with $\tw(H) \leq 3$ and for some path $P$.
\end{thm}

\cref{NewMapGraph} is an immediate consequence of \cref{d_framed_product_stucture} and the next lemma. Similar results to \cref{MewMapGraphLemma} appear in earlier works \cite{CGP06,BDGGMR,Brandenburg19,Brandenburg20} but we state the precise lemma and provide a proof here for the sake of completeness.

\begin{lem}
\label{MewMapGraphLemma}
For all integers $g\geq 0$ and $d\geq 3$, every $(g,d)$-map graph is a spanning subgraph of a $(g,d)$-framed graph.
\end{lem}

\begin{proof}
Let $G_0$ be a graph embedded in a surface of Euler genus $g$, with each face labelled a nation or a face, and where each vertex of $G_0$ is incident with at most $d$ nations. If two lakes have a common edge in their boundaries, then we may delete the edges and merge the lakes. Now assume that no two lakes have a common edge in their boundaries.
Let $G$ be the corresponding map graph.
Let $G_1$ be the graph with $V(G_1):=V(G)$, which is the set of nations.
Consider vertices $v$ and $w$ of $G_1$.
Let $A$ and $B$ be the nation faces of $G_0$ corresponding to $v$ and $w$.
Then $vw\in E(G_1)$ if $A$ and $B$ have a common edge in their boundaries,
or there is a vertex $x$ of $G_0$ in the boundary of $A$ and $B$, and in the clockwise or anticlockwise order of faces around $x$, all the faces between $A$ and $B$ are lakes.
Then $G_1$ has Euler genus at most $g$.
The faces of $G_1$ correspond to vertices of $G_0$ and lakes of $G_0$.
Triangulate each face of $G_1$ corresponding to a lake of $G_0$. For each vertex
$x$ of $G_0$ the size of the face of $G_1$ corresponding to $x$ equals the number of nations incident to $x$, which is at most $d$.
So every face of $G_1$ has size at most $d$.
If $vw\in E(G)$ then the nation faces of $G_0$ corresponding to $v$ and $w$ have a vertex $x$ of $G_0$ in common, implying that  $v$ and $w$ are both on the face of $G_1$ corresponding to $x$, and $vw\in E(\widehat{G_1})$. Hence $G\subseteq \widehat{G_1}$.
\end{proof}

\note{DW}{give a better version of \cref{MapPartition} with $g=0$. }

@@

The following corollary summarizes the results for other graph classes.
\begin{cor}\label{q_cor}
  The following bounds hold for the queue number of any graph from each of the following classes:
  \begin{compactenum}
    \item $k$-planar: $2^{O(k^3)}$
    \item $(g,k)$-planar: $g2^{O(k^3)}$
    \item $(g,\delta)$-string: $g2^{O(\delta^3)}$
    \item $1$-planar: $115$
    \item $d$-map: $24 (d+3\floor{d/2}-3) \,  - \ceil{\tfrac{3}{2} (d+3\floor{d/2}-3)}$
    \item $(g,d)$-map: $g2^{O(d)}$
    \item $k$-NN: $2^{O(k^6)}$
    % \item $(g,d)$-map graph
    % \item
  \end{compactenum}
\end{cor}

@@

The following corollary summarizes the results for other graph classes.
\begin{cor}\label{q_cor}
  The following bounds hold for the nonrepetitive chromatic number of any graph from each of the following classes:
  \begin{compactenum}
    \item $k$-planar: $(18k^2+48k+30) 4^{\binom{k+4}{3}}$
    \item $(g,k)$-planar: $\max\{2g,3\}\cdot(6k^2+16k+10) 4^{\binom{k+4}{3}}$
    \item $(g,\delta)$-string: $g4^{O(\delta^3)}$
    \item $1$-planar: $1792$
    \item $d$-map: $24 (d+3\floor{d/2}-3) \,  - \ceil{\tfrac{3}{2} (d+3\floor{d/2}-3)}$
    \item $(g,d)$-map: $g4^{O(d)}$
    \item $k$-NN: $4^{O(k^6)}$
    % \item $(g,d)$-map graph
    % \item
  \end{compactenum}
\end{cor}

Prior to the current work, the strongest upper bound on the non-repetitive chromatic number of $n$-vertex  $k$-planar graphs was $O(k\log n)$ \cite{dujmovic.morin.ea:layered}.

@@

\cref{p-centered,kPlanarProduct,1-planar,gkPlanarProduct} immediately imply the following results, for every $p\ge 2$:

\begin{cor}\quad
\begin{compactitem}
\item For every $1$-planar graph $G$,\; $ \chi_p(G)\le 5 (p+3)(p+2)(p+1)^2$.
%\item For every $k$-planar graph $G$,\; $\displaystyle \chi_p(G)\le (18k^2+48k+30)(p+1) \binom{p+ \binom{k+4}{3}-1}{ \binom{k+4}{3}-1}$.
\item For every $(g,k)$-planar graph $G$,\;
$\displaystyle \chi_p(G) \le \max\{2g,3\}\cdot(6k^2+16k+10) (p+1) \binom{p+\binom{k+4}{3}-1}{\binom{k+4}{3}-1}$.
% \note{DW}{I have changed $\binom{k+5}{4}$ to $\binom{k+4}{3}$, okay?}
% \note{PM}{Ok}
 \end{compactitem}
\end{cor}

@@

@@

\begin{thm}[\citep{DEJGMM21}]
	\label{Universal}
	For every fixed integer $t$ and every integer $n>0$ there exists a
	graph $U_n$ with $n^{1+o(1)}$ vertices such that for every graph $H$ of
	treewidth at most $t$ and path $P$, every $n$-vertex subgraph of $H\boxtimes P$ is isomorphic to an induced subgraph of $U_n$.
\end{thm}

Combining this with \cref{gkPlanarProduct,k-nn,PowerMinor,StringPartition,MapPartition} yields the following:

\begin{thm}
	\label{UniversalUniversal}
	For every fixed graph $X$ and all fixed integers $d,\delta,\Delta,g,k>0$, and every integer
	$n>0$, there exists a graph $U_n$ with $n^{1+o(1)}$ vertices such that
	$U_n$ contains all of the following graphs as induced subgraphs:
	\begin{compactitem}
		\item every $n$-vertex $(g,k)$-planar graph;
		\item every $n$-vertex $(g,d)$-map graph;
		\item every $n$-vertex $(g,\delta)$-string graph;
		\item every $n$-vertex graph $G^k$ where $G$ is $X$-minor-free and has maximum degree at most $\Delta$;
		\item every $k$-nearest neighbour graph of $n$ points in $\R^2$.
	\end{compactitem}
\end{thm}

@@

\section{Orphaned Section on Structure}

@@

\section{Orphaned Section}

\cref{nearest-neighbour,kPlanarProduct,qn,p-centered} imply:

\begin{cor}
\label{k-nn}
For every integer $k\geq 1$ there exists integers $t\leq O(k^6)$ and $\ell\leq O(k^4)$ such that every $k$-nearest-neighbour graph:
\begin{compactitem}
\item is contained in $H\boxtimes P \boxtimes K_\ell$ for some graph $H$ with treewidth $t$ and some path $P$,
\item has queue-number at most $2^{O(k^6)}$, and
\item has $p$-centered chromatic number at most $\ell (p+1)\binom{p+t}{t}$.
\end{compactitem}
\end{cor}

\cref{non-repetitive,k-nn} also give bounds on the non-repetitive chromatic number of a $k$-nearest neighbour graph $G$. However, the bound is weak, since $G$ has maximum degree at most $6k$, implying that $\pi(G) \leq (36+o(1))k^2$ by a result of \citet{DJKW16}.

\section{Orphaned Subsection}

>>>>

% \citet{DJMMUW20} introduced the study of partitions with bounded layered width such that the quotient has some additional desirable property, like small treewidth. Dujmovi\'c~et~al.\ define a class $\mathcal{G}$ of graphs to \defin{admit bounded layered partitions} if there exist $t,\ell\in\mathbb{N}$ such that every graph $G\in \mathcal{G}$ has an $H$-partition of layered width at most $\ell$ for some graph $H=H(G)$ of treewidth at most $t$.

@@

The following result is the main technical contribution of the paper.   Loosely speaking, it shows that if a graph $G$ has product structure, then so too does $G^\SS$ for every shortcut system $\SS$ of $G$.

@@

Recall that the \defin{$k$-th power} of a graph $G$ is the graph $G^k$ with vertex set $V(G^k):=V(G)$, where $vw\in E(G^k)$ if and only if $\dist_G(v,w)\leq k$. If $G$ is planar with maximum degree $\Delta$, then $G^k$ is $2k(k+1)\Delta^{k}$-planar by \cref{DrawG}.  Thus we can immediately conclude that bounded powers of planar graphs of bounded degree have product structure. However, the bounds we obtain are improved by the following lemma that constructs a shortcut system directly.

@@

% \note{PM}{Check David's commit: a0e20aff83d12c74127331854ce9fb75e002efe3}

% \note{DW}{give a better version of \cref{MapPartition} with $g=0$. }

\cref{ShortcutProduct,MapShortcut,GenusProduct} imply:

\begin{thm}
\label{gdMapProduct}
For integers $g\geq 0$ and $d\geq 3$, let $\ell:= 14\max\{2g,3\}\,\binom{d}{2}$.
Then every $(g,d)$-string graph is contained in $H\boxtimes P \boxtimes K_{\ell}$ for some path $P$ and for some graph $H$ with treewidth $9$,
\end{thm}

@@

\section{Framed Graphs, $1$-Planar Graphs, and (Plane) Map Graphs}
\label{FramedSection}

@@

\subsection{Plane Map Graphs}

% \note{PM}{Check David's commit: a0e20aff83d12c74127331854ce9fb75e002efe3}

% Map graphs are defined as follows. Start with a graph $G_0$ embedded in a surface of Euler genus $g$, with each face labelled a `nation' or a `lake', where each vertex of $G_0$ is incident with at most $d$ nations. Let $G$ be the graph whose vertices are the nations of $G_0$, where two vertices are adjacent in $G$ if the corresponding faces in $G_0$ share a vertex. Then $G$ is called a \defin{$(g,d)$-map graph}.  A $(0,d)$-map graph is called a (plane) \defin{$d$-map graph}; see \citep{FLS-SODA12,CGP02} for example. The $(g,3)$-map graphs are precisely the graphs of Euler genus at most $g$; see \citep{dujmovic.eppstein.ea:structure}.
% 
% \note{DW}{This subsection needs a clean-up; define $(g,d)$-framed graph.}

In addition to $1$-planar graphs, framed graphs can be used to obtain the following improved product structure theorem for (plane) $d$-map graphs.

\begin{thm}
\label{dMapProduct}
Every $d$-map graph is contained in $H \boxtimes P \boxtimes K_{ d + 3\floor{d/2} -3 }$ for some planar graph $H$ with $\tw(H) \leq 3$ and for some path $P$.
\end{thm}

\cref{dMapProduct} is an immediate consequence of \cref{d_framed_product_stucture} and the next lemma. Similar results to \cref{MewMapGraphLemma} appear in earlier works \cite{CGP06,BDGGMR,Brandenburg19,Brandenburg20} but we state the precise lemma and provide a proof here for the sake of completeness.

\begin{lem}
\label{MewMapGraphLemma}
For all integers $g\geq 0$ and $d\geq 3$, every $d$-map graph is a spanning subgraph of a $d$-framed graph.\todo{All I did here was remove references to $g$}
\end{lem}

\begin{proof}
\note{PM}{I fixed some issues here that come from the fact that some vertices of $G_0$ do not contribute faces to $G_1$.  This happens when two nations and two lakes meet at a vertex $x$.  It's not an issue but it means we can't write ``the face of $G_1$ that corresponds to $x$.''}
Let $G_0$ be a graph embedded in the plane, with each face labelled a nation or a face, and where each vertex of $G_0$ is incident with at most $d$ nations. If two lakes have a common edge in their boundaries, then we may delete the edges and merge the lakes. Now assume that no two lakes have a common edge in their boundaries.
Let $G$ be the corresponding map graph.
Let $G_1$ be the graph with $V(G_1):=V(G)$, which is the set of nations, and whose edge set we now define.  Consider vertices $v$ and $w$ of $G_1$.
Let $A$ and $B$ be the nation faces of $G_0$ corresponding to $v$ and $w$.
Then $vw\in E(G_1)$ if $A$ and $B$ have a common edge in their boundaries,
or there is a vertex $x$ of $G_0$ in the boundary of $A$ and $B$, and in the clockwise or anticlockwise order of faces around $x$, all the faces between $A$ and $B$ are lakes.
% Then $G_1$ has Euler genus at most $g$.
Each face of $G_1$ corresponds to a vertex of $G_0$ or a lake of $G_0$.
Triangulate each face of $G_1$ corresponding to a lake of $G_0$.

For each face $F$ of $G_1$ that corresponds to a vertex $x$ of $G_0$, the size of $F$ equals the number of nations incident to $x$, which is at most $d$.
So every face of $G_1$ has size at most $d$.  Let $\hat{G}_1$ be the $d$-framed graph whose frame is $G_1$.

To complete the proof, we now show that $\hat{G}_1$ contains $G$.
If $vw\in E(G)$ then the nation faces of $G_0$ corresponding to $v$ and $w$ have a vertex $x$ of $G_0$ in common.   If some face of $G_1$ corresponds to $x$ then $v$ and $w$ are on a common face of $G_1$ so $vw\in E(\hat{G}_1)$.  If no face of $G_1$ corresponds to $x$ then the only two nation faces of $G_0$ with $x$ on their boundary those corresponding to $v$ and $w$, in which case $vw\in E(G_1)\subseteq E(\hat{G}_1)$.  Therefore $G\subseteq\hat{G}_1$.
\end{proof}

% \note{DW}{give a better version of \cref{MapPartition} with $g=0$. }

@@

The following corollary summarizes the results for other graph classes.
\begin{cor}\label{q_cor}
  The following bounds hold for the queue number of any graph from each of the following classes:
  \begin{compactenum}
    \item $k$-planar: $2^{O(k^3)}$
    \item $(g,k)$-planar: $\max\{1,g\}\cdot 2^{O(k^3)}$
    \item $(g,\delta)$-string: $\max\{1,g\}\cdot 2^{O(\delta^3)}$
    \item $1$-planar: $115$
    \item $d$-map: $24 (d+3\floor{d/2}-3) \,  - \ceil{\tfrac{3}{2} (d+3\floor{d/2}-3)}$
    \item $(g,d)$-map: $\max\{1,g\}\cdot 2^{O(d)}$
    \item $k$-NN: $2^{O(k^6)}$
    % \item $(g,d)$-map graph
    % \item
  \end{compactenum}
\end{cor}

@@

Combining \cref{non-repetitive} with our results on product structure, we obtain the following corollary:
\begin{cor}\label{nonrepetitive_cor}
  The following bounds hold for the nonrepetitive chromatic number of any graph from each of the following classes:
  \begin{compactenum}
    \item $k$-planar: $(18k^2+48k+30)\cdot 4^{\binom{k+4}{3}}$
    \item $(g,k)$-planar: $\max\{2g,3\}\cdot(6k^2+16k+10) 4^{\binom{k+4}{3}}$
    % \item $(g,\delta)$-string: $g4^{O(\delta^3)}$
    \item $1$-planar: $1792$
    \item $d$-map: $256(d+3\floor{d/2}-3)$
    % \item $(g,d)$-map: $g4^{O(d)}$
    % \item $k$-NN: $4^{O(k^6)}$
    \item $(g,d)$-map graph: \textcolor{red}{??}
    % \item
  \end{compactenum}
\end{cor}

Prior to the current work, the strongest upper bound on the non-repetitive chromatic number of $n$-vertex  $k$-planar graphs was $O(k\log n)$ \cite{dujmovic.morin.ea:layered}.  Our results also give bounds on the nonrepetitive chromatic number of $(g,\delta)$-string graphs and $k$-nearest-neighbour graphs, but these bounds are weaker than existing results based on maximum degree.  The bounds obtained from product structure are exponential in $\delta$ and $k^2$, respectively.  However, each of these graph classes has degree bounded by $O(\delta)$ and $O(k^2)$, respectively, and therefore have nonrepetitive chromatic number $O(\delta^2)$ and $O(k^{4})$, respectively \cite{DJKW16}.

@@

% \cref{p-centered,kPlanarProduct,1-planar,gkPlanarProduct} immediately imply the following results, for every $p\ge 2$:
% 
% \begin{cor}\quad
% \begin{compactitem}
% \item For every $1$-planar graph $G$,\; $ \chi_p(G)\le 5 (p+3)(p+2)(p+1)^2$.
% %\item For every $k$-planar graph $G$,\; $\displaystyle \chi_p(G)\le (18k^2+48k+30)(p+1) \binom{p+ \binom{k+4}{3}-1}{ \binom{k+4}{3}-1}$.
% \item For every $(g,k)$-planar graph $G$,\;
% $\displaystyle \chi_p(G) \le \max\{2g,3\}\cdot(6k^2+16k+10) (p+1) \binom{p+\binom{k+4}{3}-1}{\binom{k+4}{3}-1}$.
% % \note{DW}{I have changed $\binom{k+5}{4}$ to $\binom{k+4}{3}$, okay?}
% % \note{PM}{Ok}
%  \end{compactitem}
% \end{cor}

For planar graphs of treewidth at most $3$ an $O(p^2\log p)$ upper bound is known:

\begin{lem}\cite{DFMS21}\label{simple_3tree_centered_colouring}
  For every planar graph $H$ of treewidth at most $3$, 
  \[ 
    \chi_p(H)\le (p+1)(p\ceil{\log(p+1)}+2p+1) \enspace .
  \]
\end{lem}

Combining \cref{p-centered,simple_3tree_centered_colouring} with our results on product structure, we obtain the following corollary:

\begin{cor}\label{p_centered_cor}
  The following bounds hold for the $p$-centered chromatic number of any graph from each of the following classes:
  \begin{compactenum}
    \item $k$-planar: $(18k^2+48k+30)\cdot 4^{\binom{k+4}{3}}$
    \item $(g,k)$-planar: $\max\{2g,3\}\cdot(6k^2+16k+10)(p+1)\binom{p+\binom{k+4}{3}-1}{\binom{k+4}{3}-1}$
    % \item $(g,\delta)$-string: $g4^{O(\delta^3)}$
    \item $1$-planar: $7(p+1)^2(p\ceil{\log(p+1)}+2p+1)$
    \item $d$-map: $(d+3\floor{d/2}-3)\cdot (p+1)^2(p\ceil{\log(p+1)}+2p+1)$
    % \item $(g,d)$-map: $g4^{O(d)}$
    % \item $k$-NN: $4^{O(k^6)}$
    \item $(g,d)$-map: $\max\{2g,3\}\cdot 7d(d-3)(p+1)\cdot \binom{p+9}{9}$
    % \item
  \end{compactenum}
\end{cor}

@@

As with nonrepetitive chromatic number, our results also imply bounds on the $p$-centered chromatic numer of $(g,\delta)$-string graphs and $k$-nearest-neighbour graphs, but these bounds are weaker than the bounds implied by the fact that these graphs have bounded maximum degree \cite{DFMS21}.

@@

\begin{thm}[\citep{DEJGMM21,EJM}]
	\label{Universal}
	For every fixed integer $t$ and every integer $n>0$ there exists a
	graph $U_n$ with $n^{1+o(1)}$ vertices and edges such that for every graph $H$ of treewidth at most $t$ and path $P$, every $n$-vertex subgraph of $H\boxtimes P$ is isomorphic to an induced subgraph of $U_n$.
\end{thm}

Combining \cref{Universal} with our results on product structure yields the following:
% this with \cref{gkPlanarProduct,k-nn,PowerMinor,StringPartition,MapPartition} yields the following:

\begin{cor}\label{universal_cor}
	\label{UniversalUniversal}
	For every fixed graph $X$ and all fixed integers $d,\delta,\Delta,g,k>0$, and every integer $n>0$, there exists a graph $U_n$ with $n^{1+o(1)}$ vertices and edges such that $U_n$ contains all of the following graphs as induced subgraphs:
	\begin{compactitem}
		\item every $n$-vertex $(g,k)$-planar graph;
		\item every $n$-vertex $(g,d)$-map graph;
		\item every $n$-vertex $(g,\delta)$-string graph;
		\item every $n$-vertex graph $G^k$ where $G$ is $X$-minor-free and has maximum degree at most $\Delta$;
		\item every $k$-nearest neighbour graph of $n$ points in $\R^2$.
	\end{compactitem}
\end{cor}

@@

\section{Orphaned Text}

\subsection{Text following \cref{gkPlanarProduct}}

@@

% \subsection{Orphaned Section}
% 
% \cref{nearest-neighbour,kPlanarProduct,qn,p-centered} imply:
% 
% \begin{cor}
% \label{k-nn}
% For every integer $k\geq 1$ there exists integers $t\leq O(k^6)$ and $\ell\leq O(k^4)$ such that every $k$-nearest-neighbour graph:
% \begin{compactitem}
% \item is contained in $H\boxtimes P \boxtimes K_\ell$ for some graph $H$ with treewidth $t$ and some path $P$,
% \item has queue-number at most $2^{O(k^6)}$, and
% \item has $p$-centered chromatic number at most $\ell (p+1)\binom{p+t}{t}$.
% \end{compactitem}
% \end{cor}
% 
% \cref{non-repetitive,k-nn} also give bounds on the non-repetitive chromatic number of a $k$-nearest neighbour graph $G$. However, the bound is weak, since $G$ has maximum degree at most $6k$, implying that $\pi(G) \leq (36+o(1))k^2$ by a result of \citet{DJKW16}.
